### 2020-10-19 tex/initial-presentation.tex
Initial commit.

\documentclass[aspectratio=169]{beamer}

\usetheme{metropolis}

\usefonttheme{professionalfonts}
\usepackage[familydefault,light]{Chivo}

\input{common.tex}

\date{\formatdate{20}{10}{2020}}

\institute{supervised by Prof. Dr. Ruth Breu}

\begin{document}
  \maketitle

  \begin{frame}{Outline}
    \begin{itemize}
      \item Motivation
      \item Low-Code Architecture
      \item Model-Driven Architecture
      \item Criticisms
      \item Planned Project Procedure
    \end{itemize}
  \end{frame}

  \begin{frame}{Motivation}
    \begin{itemize}
      \item What can and can't low-code and model-driven tools do?
      \item Are they a viable alternative to traditional development?
      \item When to choose one approach over the other?
    \end{itemize}
  \end{frame}

  \begin{frame}{Low-Code Architecture}
    \begin{itemize}
      \item pre-built application components
      \item graphical user interface for creating application logic as well as
    \end{itemize}
  \end{frame}

  \begin{frame}{Model-Driven Architecture}
    \begin{itemize}
      \item
    \end{itemize}
  \end{frame}

  \begin{frame}{Criticisms}
    \begin{itemize}
      \item Unsuitable for implementing scalable and mission-critical applications.
      \item Do these approaches actually make development easier and cheaper?
      \item Increase in unsupported applications built by shadow IT.
    \end{itemize}
  \end{frame}

  \begin{frame}{Planned Project Procedure}
    \begin{itemize}
      \item evaluate Low-Code tools
      \item evaluate Model-Driven tools
      \item pick two to three of each
      \item compare the two sets with each other
      \item check if a tool fulfils
    \end{itemize}
  \end{frame}
\end{document}

### 2020-10-19 tex/paper.tex
Initial commit.

\documentclass[runningheads]{llncs}

\linespread{1.1}

\raggedbottom

\usepackage{float}

\input{common.tex}

\PassOptionsToPackage{hyphens}{url}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{nameref}

\addto\extrasenglish{
  \def\sectionautorefname{Section}
}

\usepackage[
  style   = numeric,
  sorting = none,
]{biblatex}
\bibliography{paper}

\begin{document}

\institute{University of Innsbruck, Austria}

\maketitle

\vspace{8em}

\begin{abstract}

In this paper we present a survey conducted by comparing low-code tools with model-driven tools for developing software in order to get a better understanding of the advantages and drawbacks of both approaches and whether they are a viable alternative to traditional software development.

\keywords{Low-Code \and Model-Driven \and Development}

\end{abstract}

\newpage

\section{Introduction}
\label{sec:introduction}

\section{Evaluation of Low-Code Tools}
\label{sec:evaluation_of_low_code_tools}

\section{Evaluation of Model-Driven Tools}
\label{sec:evaluation_of_model_driven_tools}

\section{Findings}
\label{sec:findings}

\section{Conclusion}
\label{sec:conclusion}

\newpage
\printbibliography

\end{document}

### 2020-10-19 tex/initial-presentation.tex
Complete initial presentation.

@@
     \begin{itemize}
-      \item pre-built application components
-      \item graphical user interface for creating application logic as well as
+      \item provides pre-built application components
+      \item graphical user interface for creating both the
+            application logic as well as the user interface
+      \item typically aimed at end-users rather than developers
     \end{itemize}
@@
     \begin{itemize}
-      \item
+      \item provides a set of guidelines for the structuring of specifications
+      \item code (fully or partially) generated from models, e.g. from UML diagrams
+      \item aimed at developers to speed up development
     \end{itemize}
@@
     \begin{itemize}
+      \item Low-Code Architecture
+        \begin{itemize}
           \item Unsuitable for implementing scalable and mission-critical applications.
+          \item Increase in unsupported applications built by shadow IT,
+                i.e. applications which are not controlled by a company's IT department.
+        \end{itemize}
+      \item Model-Driven Architecture
+        \begin{itemize}
+          \item UML diagrams lack details included in the code itself.
+          \item “the Code is the design” - Should models be derived from code instead of code from models?
+        \end{itemize}
       \item Do these approaches actually make development easier and cheaper?
-      \item Increase in unsupported applications built by shadow IT.
     \end{itemize}
@@
     \begin{itemize}
-      \item evaluate Low-Code tools
-      \item evaluate Model-Driven tools
-      \item pick two to three of each
-      \item compare the two sets with each other
-      \item check if a tool fulfils
+      \item Evaluate a set of Low-Code tools.
+      \item Evaluate a set of Model-Driven tools.
+      \item For each tool, determine what it can and can't do.
+      \item Compare the two sets with each other and draw some conclusions.
     \end{itemize}

### 2020-12-27 tex/paper.tex
Add “What is …?” sections.

@@
 
+In this paper, we compare Low-Code Platforms to Model-Driven Architecture and try to determine when and where to use them over Model-Driven Architecture or vice-versa.
+
+In order to make clear to the reader the difference between the two approaches we firstly provide definitions and characteristics of each paradigm in the following sections.
+
+\subsection{What is Model-Driven Architecture?}
+\label{ssec:what_is_model_driven}
+
+With Model-Driven Architecture or Model-Driven Engineering, the goal is to standardise on models in a given domain.
+This is done in order to reduce code duplication and speed up development.
+Especially when bootstrapping a new project, it is helpful keep boilerplate code to a minimum.
+In addition to reducing development time, Model-Driven Engineering also tries to increase productivity by maximising compatibility between systems by using the aforementioned standardised models.
+\cite{wiki:model_driven_engineering} Generally, Model-Driven Architectures are geared towards developers, i.e.
+people who also have a good understanding of the underlying programming language(s) of the respective architecture.
+One common way Model-Driven Engineering is used in practice is code generation.
+Using code generation can speed up many processes: For example, given an API specification (e.g.
+using the industry standard OpenAPI \cite{openapis}), one can generate all code that is necessary for a client for this API.
+Swagger \cite{swagger} is one such tool which can generate API clients for many different programming languages.
+There are also tools which can generate a complete class hierarchy from an UML diagram; many IDEs offer such functionality either natively or via plug-ins.
+These are only a few examples of what falls into the category of Model-Driven Architecture, the main point is that all of them help developers reach their goal faster than doing the same tasks manually.
+
+\subsection{What is a Low-Code Platform?}
+\label{ssec:what_is_a_low_code_platform}
+
+A low-code platform, as the name implies, requires a minimal amount of code to be written.
+Similarly to conventional programming (i.e.
+using a programming language), an IDE is used for creating programs using low-code platforms.
+Often, the IDE is part of the low-code platform itself instead of a standalone program as is the case for conventional programming.
+In most cases, the IDE of the platform and the low-code platform cannot be used separately so the term low-code platform is also commonly used to refer to the IDE interface itself.
+Low-code platforms are targeted towards users with little to no programming experience, which means the actual programming of a low-code system is done mainly via visual building blocks.
+These building blocks contain pre-built templates and functionality and can be combined to form a coherent user interface.
+Oftentimes, custom functionality can be implemented using a click-and-drag pattern between building blocks and by assigning an action accordingly.
+The available building blocks vary by platform which makes them less portable than a system written using a programming language.
+This is one common point of critique on low-code platforms.
+
 \section{Evaluation of Low-Code Tools}

### 2021-01-02 tex/paper.tex
Extend paper.

@@
 Especially when bootstrapping a new project, it is helpful keep boilerplate code to a minimum.
-In addition to reducing development time, Model-Driven Engineering also tries to increase productivity by maximising compatibility between systems by using the aforementioned standardised models.
-\cite{wiki:model_driven_engineering} Generally, Model-Driven Architectures are geared towards developers, i.e.
+In addition to reducing development time, Model-Driven Engineering also tries to increase productivity by maximising compatibility between systems by using the aforementioned standardised models.~\cite{wiki:model_driven_engineering} Generally, Model-Driven Architectures are geared towards developers, i.e.
 people who also have a good understanding of the underlying programming language(s) of the respective architecture.
@@
 Using code generation can speed up many processes: For example, given an API specification (e.g.
-using the industry standard OpenAPI \cite{openapis}), one can generate all code that is necessary for a client for this API.
-Swagger \cite{swagger} is one such tool which can generate API clients for many different programming languages.
+using the industry standard OpenAPI~\cite{openapis}), one can generate all code that is necessary for a client for this API.
+Swagger~\cite{swagger} is one such tool which can generate API clients for many different programming languages.
 There are also tools which can generate a complete class hierarchy from an UML diagram; many IDEs offer such functionality either natively or via plug-ins.
@@
 
+Another sub-category of low-code platforms are so called no-code platforms~\cite{wiki:no_code_development_platform}.
+Where to draw the line between low-code and no-code is not clearly defined, especially since low-code platforms can sometimes be used in a way that cannot be differentiated from no-code platforms.
+For the purposes of this paper, low-code platforms are general purpose visual programming languages~\cite{wiki:visual_programming_language} aimed at both developer and non-developer users while no-code platforms are closer to software as a service (SaaS) platforms targeted mostly (or only) towards non-developers.
+Another way to look at it is that low-code platforms still give developers the freedom to add custom building blocks for their specific use case and no-code solutions are usually purpose-made for a specific line of business or a specific type of department.
+Some examples include monday.com~\cite{wiki:monday_com}, a no-code platform for project management tasks; FileMaker~\cite{wiki:filemaker}, which provides a database engine and a corresponding no-code platform for creating graphical user interfaces for interacting with databases; and Airtable~\cite{wiki:airtable}, a no-code platform for creating databases connected to interactive spreadsheets.
+
 \section{Evaluation of Low-Code Tools}
 \label{sec:evaluation_of_low_code_tools}
+
+\subsection{Selection of Low-Code Tools}
+\label{subsec:choosing_low_code_tools}
+
+Before evaluating low-code tools, we had to first determine which tools to evaluate.
+Given the ever evolving landscape of low-code platforms, there are countless companies offering low-code solutions.
+The following is an incomplete list of low-code platforms we found:
+
+\begin{itemize}
+  \item Oracle APEX (Application Express) (originally released as Oracle Flows in 2000) by Oracle (founded in 1977)
+  \item OutSystems (founded in 2001)
+  \item Mendix (founded in 2005), a subsidiary of Siemens
+  \item Simplifier (founded in 2012 as iTiZZiMO, renamed 2019 to Simplifier, the same name as its low-code platform)
+  \item OSBP (Open Standard Buiness Platform), an open-source low-code platform (developed since 2016) under the umbrella of the Eclipse Foundation
+  \item Corteza Low Code, part of the open-source Corteza Project (released in 2019), originally developed by Crust Technology
+\end{itemize}
+
+We wanted our evaluation to contain at least one open-source platform (e.g.
+OSBP), one platform backed by a well-known corporation (e.g.
+Oracle), one platform by a relatively new/unknown corporation (e.g.
+Simplifier), one relatively old/well-established platform (e.g.
+Oracle APEX or OutSystems) and one relatively new platform (e.g.
+Simplifier or Corteza).
+With the given list of low-code platforms, we could satisfy this requirement.
 

### 2021-01-02 tex/paper.tex
Add initial sections about OSBP and Corteza.

@@
 
+To ensure we could properly test all of these tools, we conducted a brief first inspection of each.
+Therefore, the following section will give a brief explanation of each platform, any problems encountered during the initial setup and the steps necessary before being able to use the low-code platform.
+
+\subsubsection{OSBP (Open Standard Business Platform)}
+
+The Open Standard Business Platform is an open-source development environment which combines no-code and low-code with traditional application development.~\cite{osbp} It is the community version of OS.bee~\cite{osbee}, a commercial product developed by COMPEX only available to business users.
+The interface for the OSBP platform is implemented via a plug-in for the Eclipse IDE.
+The user first needs to install the Eclipse IDE, after which they can add the link to the plug-in repository and install it directly from within the IDE.
+Unfortunately, the latest version of OSBP was released more than one year ago and is incompatible with the latest version of the Eclipse IDE.
+At the time of writing (Jan 2, 2021), the last commit in the OBSP Git repository was on May 9, 2019.
+The latest version of the Eclipse IDE was 2020-12.
+The OBSP website recommended downloading Eclipse Neon, a version of the Eclipse IDE released in 2016.
+A non-functioning IDE plug-in for an outdated version of said IDE which has not been updated in over a year is a red flag.
+We therefore deemed OSBP unsuitable and decided to exclude it from further evaluation.
+
+\subsubsection{Corteza}
+
+Corteza Low Code is part of the Corteza Project~\cite{corteza} initiated by Curst Technology.
+The Corteza Project encompasses Corteza CRM Suite, an open-source customer relationship management tool built entirely on Corteza Low Code; Corteza Messaging, a messaging solution similar to the likes of Slack; Corteza One, a unified workspace from which to access and run web applications and finally Corteza Low Code, a low-code platform designed for building record-based applications.
+It is possible to try out Corteza directly in the browser by creating an account or by using an existing GitHub or Google account to sign up.
+We chose to set up a local demo deployment using Docker.
+The Corteza Project provides very good documentation on how to get started, including a ready-made `docker-compose.yml' file containing all necessary services for a full Corteza installation.
+After a simple call to \texttt{docker-compose up -d}, the Corteza server was running locally.
+The last step before being ready to use was to set up an administrator account in the local Corteza instance.
+
 \section{Evaluation of Model-Driven Tools}

### 2021-01-02 tex/paper.tex
Fix spelling.

@@
   \item Simplifier (founded in 2012 as iTiZZiMO, renamed 2019 to Simplifier, the same name as its low-code platform)
-  \item OSBP (Open Standard Buiness Platform), an open-source low-code platform (developed since 2016) under the umbrella of the Eclipse Foundation
+  \item OSBP (Open Standard Business Platform), an open-source low-code platform (developed since 2016) under the umbrella of the Eclipse Foundation
   \item Corteza Low Code, part of the open-source Corteza Project (released in 2019), originally developed by Crust Technology
@@
 Unfortunately, the latest version of OSBP was released more than one year ago and is incompatible with the latest version of the Eclipse IDE.
-At the time of writing (Jan 2, 2021), the last commit in the OBSP Git repository was on May 9, 2019.
+At the time of writing (Jan 2, 2021), the last commit in the OSBP Git repository was on May 9, 2019.
 The latest version of the Eclipse IDE was 2020-12.
-The OBSP website recommended downloading Eclipse Neon, a version of the Eclipse IDE released in 2016.
+The OSBP website recommended downloading Eclipse Neon, a version of the Eclipse IDE released in 2016.
 A non-functioning IDE plug-in for an outdated version of said IDE which has not been updated in over a year is a red flag.
@@
 
-Corteza Low Code is part of the Corteza Project~\cite{corteza} initiated by Curst Technology.
+Corteza Low Code is part of the Corteza Project~\cite{corteza} initiated by Crust Technology.
 The Corteza Project encompasses Corteza CRM Suite, an open-source customer relationship management tool built entirely on Corteza Low Code; Corteza Messaging, a messaging solution similar to the likes of Slack; Corteza One, a unified workspace from which to access and run web applications and finally Corteza Low Code, a low-code platform designed for building record-based applications.
@@
 We chose to set up a local demo deployment using Docker.
-The Corteza Project provides very good documentation on how to get started, including a ready-made `docker-compose.yml' file containing all necessary services for a full Corteza installation.
+The Corteza Project provides very good documentation on how to get started, including a ready-made \texttt{docker-compose.yml} file containing all necessary services for a full Corteza installation.
 After a simple call to \texttt{docker-compose up -d}, the Corteza server was running locally.

### 2021-01-03 tex/paper.tex
Change `cleveref` configuration.
Add implementation of Corteza application and Oracle APEX setup.

@@
 \usepackage{hyperref}
-\usepackage{cleveref}
+\usepackage[nameinlink, noabbrev]{cleveref}
 \usepackage{nameref}
@@
 
+\subsubsection{Oracle APEX (Application Express)}
+
+Oracle Application Express~\cite{oracle_apex} is a web-based application development platform developed by Oracle.
+It is backed by an Oracle database and allows building web applications using a low-code interface.
+Oracle offers three ways to set up APEX.
+The first is to use it as part of their Oracle Cloud offering, secondly, by requesting a free workspace or lastly by self-hosting using the Oracle APEX On Premise variant.
+For our purposes, we chose to request a free workspace for testing.
+We chose a name for our namespace and registered using an email address and shortly thereafter we were sent an approval email with the login details.
+After that, Oracle APEX was ready to use.
+
+\subsection{Developing a TODO List Application}
+
+In order to evaluate each platform, we decided to build a simple TODO List application.
+The user should be able to add items to the list containing a title, body text and the status of the TODO item (either “done” or “outstanding”).
+
+\subsubsection{Corteza}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.6\textwidth]{corteza-home-screen}
+  \caption{Corteza Home Screen}
+  \label{fig:corteza_home_screen}
+\end{figure}
+
+In \cref{fig:corteza_home_screen}, we can see the initial screen of the Corteza platform.
+By selecting “Low Code”, we reach a new view containing all existing “Low Code Namespaces”.
+A namespace can be thought of as an individual application within the low-code environment.
+In order to start with our application, we create a \texttt{todo-list} namespace.
+Next we need to create a module.
+A module in Corteza is similar to creating a database schema for a single table.
+For our application, we create a module containing text fields for \texttt{title} and \texttt{body} and a dropdown field for the \texttt{status}.
+We also mark \texttt{title} as a required field, as seen in \cref{fig:corteza_module}.
+Next, we need to create a page which lets us add TODOs to our list.
+To do this, we create a blank page and add the pre-built “Record List” block.
+In the “Record List” block, we select our “Todo Module” as the data source and select all fields so they are displayed in the list.
+The list also provides an option to allow inline editing of records.
+Without this option, the user has to click on the list item to open the detail page in order to edit it.
+Unfortunately, for some reason when enabling the checkbox to allow inline editing, we were unable to save the page with this configuration.
+Nevertheless, the basic functionality of our application was completely implemented in less than ten minutes.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{corteza-module}
+  \caption{Creation of a Corteza Module}
+  \label{fig:corteza_module}
+\end{figure}
+
+There are also some other features in Corteza: The “Chart Builder” allows creating charts for data in a module and it is possible add custom buttons with automations.
+The custom automations however, have to be fully implemented in JavaScript, so we didn't evaluate them any further given our focus on low-code.
+
+\subsubsection{Oracle APEX (Application Express)}
+
 \section{Evaluation of Model-Driven Tools}

### 2021-01-04 tex/paper.tex
Finish Oracle APEX section.

@@
 
-In order to evaluate each platform, we decided to build a simple TODO List application.
+In order to evaluate each platform, we decided to build a simple TODO list application.
 The user should be able to add items to the list containing a title, body text and the status of the TODO item (either “done” or “outstanding”).
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{oracle-apex-home-screen}
+  \caption{Oracle APEX Home Screen}
+  \label{fig:oracle_apex_home_screen}
+\end{figure}
+
+When first logging into Oracle APEX, the user is greeted with a an overview of all applications in the workspace and a quick menu for accessing the major parts, App Builder, SQL Workshop, Team Development and App Gallery, as seen in \cref{fig:oracle_apex_home_screen}.
+App Builder is the low-code development interface, allowing the user to create page-based web applications, SQL Workshop is an interface to create database tables for use in low-code applications.
+Team Development is an integrated project management tool that helps coordinate tasks across team members when multiple people are working together in the same namespace.
+And finally, App Gallery is a gallery containing sample applications to use as either templates or guidance when developing a new low-code application.
+
+Before starting with our TODO list application, we decided to set up a sample application called “Sample Database Application”, which contains a basic inventory management application, including a product list which we used as a template for our TODO list.
+
+For our todo list, we created a new application called “TODO List”.
+The setup wizard allows customising the color theme of the application, the menu style (e.g.
+side menu or top menu), language as well as the application icon.
+Once created, we had a blank application containing a login page and a home page.
+The next thing we needed to do was set up a database table for our todo items.
+Basic SQL knowledge is required to do this since properly setting up a primary key and a unique constraint for it is necessary to be able to insert data later on.
+We now had a \texttt{TODOS} table containing \texttt{ID}, \texttt{STATUS}, \texttt{TITLE} and \texttt{BODY} and were now able to add a new page to our application.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{oracle-apex-create-page}
+  \caption{Oracle APEX Create Page Dialog}
+  \label{fig:oracle_apex_create_page}
+\end{figure}
+
+There are many different page styles to choose from, as seen in \cref{fig:oracle_apex_create_page}.
+For our use case, we chose the “Form” page template.
+The user simply assigns a page name and subsequently selects the  corresponding database table.
+The result is a form page with all fields of the database table which lets the user add new tasks and edit existing ones.
+Some minor visual adjustments were necessary for the form page, i.e.
+hiding the automatically generated \texttt{ID} and setting the \texttt{STATUS} to “outstanding” by default when adding a new task.
+The next step was to create a list containing our todo items with a button that opens the form page to add new ones.
+Similar to the page templates, there are many different building blocks to choose from.
+We used the same as the one used in the sample application, the “Interactive Report”, a table view with filtering options.
+Lastly, we added a \texttt{CREATE} button to the top of this component, linking it to the form page created earlier.
+
+Oracle APEX provides many advanced options for all building blocks which allows building custom logic by connecting them via events (e.g.
+mouse click or value change).
+It is also possible to create custom events: For example, a form page (in dialog form) provides a \texttt{Cancel Dialog} event which is fired when the \texttt{CANCEL} button is pressed.
+
 \section{Evaluation of Model-Driven Tools}

### 2021-01-04 tex/paper.tex
Fix spelling.
Add Simplifier and Mendix setup.

@@
 
+\subsubsection{Simplifier}
+
+Simplifier is, like Oracle APEX, a web-based application development environment on low-code basis.
+Simplifier offers a demo platform called “Simplifier Playground”, an instance of the Simplifier platform specifically set up for trialling the platform.
+The trial instance is reset to defaults every night, so all data stored on it will not be persisted.
+They also offer a “Freemium” instance, available by registering via an email address, which has no such limitation, we therefore applied for such an instance.
+%TODO
+
+\subsubsection{Mendix}
+
+Similar to Oracle APEX and Simplifier, Mendix is also a web-based environment for low-code application development.
+Mendix offerse a free tier which allows hosting unlimited applications with 1GB of memory and 0.5GB of storage per application.
+For more advanced needs, Mendix offers an enterprise option starting at 1875€ per month, allowing on-premises hosting and hosting on private clouds, e.g.
+AWS, IBM etc.
+For our evaluation purposes, the free tier is more than enough given there are no restrictions on the provided functionality.
+Creating an account was straight-forward and we were able to log in within minutes of registering.
+There are two options for developing applications using Mendix: The first is the web-based environment called Mendix Studio, the second a native desktop application called Mendix Studio Pro, offering more advanced options for professional users.
+Unfortunately Mendix Studio Pro is only available for Microsoft Windows, so we were not able to test it, the basic functionality for creating low-code application is the same in both versions of Mendix Studio, however.
+
 \subsection{Developing a TODO List Application}
@@
 A module in Corteza is similar to creating a database schema for a single table.
-For our application, we create a module containing text fields for \texttt{title} and \texttt{body} and a dropdown field for the \texttt{status}.
+For our application, we create a module containing text fields for \texttt{title} and \texttt{body} and a drop-down field for the \texttt{status}.
 We also mark \texttt{title} as a required field, as seen in \cref{fig:corteza_module}.

### 2021-01-04 tex/paper.tex
Add Mendix section.

@@
 
+\subsubsection{Mendix}
+
+Mendix offerse three ways for creating a new application: Creating a blank application from scratch, generating an application from spreadsheet data or using one of many templates as a starting point.
+For our TODO list application, we chose to create an application from scratch.
+After creating an application, you can either edit it using the web interface or open it in the Mendix Studio Pro desktop application.
+We developed our TODO list application entirely using the web interface.
+The Mendix interface provides four main menu items: Pages, Domain Models, Microflows and Navigation.
+Pages is used to manage the different pages needed for the application.
+Domain Models contains the entities for our application, in our case we needed an entity for items in our TODO list.
+Microflows are used to add custom logic (micro workflows) to the application, e.g.
+when clicking a button.
+An interesting feature is Mendix Assist, an AI-based assistant making suggestions to the user when creating Microflows.
+Lastly we have Navigation, which is simply the representation of the application menu hierarchy.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\textwidth]{mendix-create-page}
+  \caption{Mendix Create Page Dialog}
+  \label{fig:mendix_create_page}
+\end{figure}
+
+For our application, we started by adding a new page for creating a task on our TODO list.
+There are numerous page templates to choose from, we used the “Form Horizontal” template as seen in \cref{fig:mendix_create_page}.
+The resulting page (\texttt{add\_task}) contains a so called data view.
+Every data view needs a data source, so we were prompted to create a new entity.
+We added a new entity called \texttt{task} and while editing the form, we were prompted to add attributes to that entity.
+We added two \texttt{String} attributes for \texttt{title} and \texttt{body}.
+We also needed to add a \texttt{status} attribute; for this, we edited the entity directly, adding an attribute of type \texttt{Enumeration} with values for “done” and “outstanding” and set the default value to “outstanding” accordingly.
+
+Next, we added a list view to the \texttt{home} page, choosing the \texttt{task} entity as data source in the process.
+We could now drag text fields into the list view and choose the corresponding attributes to be displayed.
+In order to actually add new tasks, we also added a “New task” button.
+For the click action, the user can choose to open a page or run a microflow or a predefined action such as saving changes on the current page.
+We chose our page for creating a new task as the target and also needed to enable the “Create object” option.
+With this option, we pass an empty \texttt{task} object to the page, since the data view on the page expects a \texttt{task} as data source.
+
+Lastly, we created another form page (\texttt{edit\_task}), also containing the \texttt{status}, in order to be able to edit a \texttt{task}.
+To access it, we added a click event to the list view on the \texttt{home} page.
+Since the list view data source is a \texttt{task}, it is automatically passed to the data view on the \texttt{edit\_task} page.
+
 \section{Evaluation of Model-Driven Tools}

### 2021-01-04 tex/paper.tex
Fix spelling.

@@
 Similar to Oracle APEX and Simplifier, Mendix is also a web-based environment for low-code application development.
-Mendix offerse a free tier which allows hosting unlimited applications with 1GB of memory and 0.5GB of storage per application.
+Mendix offers a free tier which allows hosting unlimited applications with 1GB of memory and 0.5GB of storage per application.
 For more advanced needs, Mendix offers an enterprise option starting at 1875€ per month, allowing on-premises hosting and hosting on private clouds, e.g.
@@
 To do this, we create a blank page and add the pre-built “Record List” block.
-In the “Record List” block, we select our “Todo Module” as the data source and select all fields so they are displayed in the list.
+In the “Record List” block, we select our “TODO Module” as the data source and select all fields so they are displayed in the list.
 The list also provides an option to allow inline editing of records.
@@
 
-For our todo list, we created a new application called “TODO List”.
-The setup wizard allows customising the color theme of the application, the menu style (e.g.
+For our TODO list, we created a new application called “TODO List”.
+The setup wizard allows customising the colour theme of the application, the menu style (e.g.
 side menu or top menu), language as well as the application icon.
 Once created, we had a blank application containing a login page and a home page.
-The next thing we needed to do was set up a database table for our todo items.
+The next thing we needed to do was set up a database table for our TODO items.
 Basic SQL knowledge is required to do this since properly setting up a primary key and a unique constraint for it is necessary to be able to insert data later on.
@@
 hiding the automatically generated \texttt{ID} and setting the \texttt{STATUS} to “outstanding” by default when adding a new task.
-The next step was to create a list containing our todo items with a button that opens the form page to add new ones.
+The next step was to create a list containing our TODO items with a button that opens the form page to add new ones.
 Similar to the page templates, there are many different building blocks to choose from.

### 2021-01-04 tex/paper.tex
Fix spelling.

@@
 
-Mendix offerse three ways for creating a new application: Creating a blank application from scratch, generating an application from spreadsheet data or using one of many templates as a starting point.
+Mendix offers three ways for creating a new application: Creating a blank application from scratch, generating an application from spreadsheet data or using one of many templates as a starting point.
 For our TODO list application, we chose to create an application from scratch.
@@
 In order to actually add new tasks, we also added a “New task” button.
-For the click action, the user can choose to open a page or run a microflow or a predefined action such as saving changes on the current page.
+For the click action, the user can choose to open a page or run a Microflow or a predefined action such as saving changes on the current page.
 We chose our page for creating a new task as the target and also needed to enable the “Create object” option.

### 2021-01-09 tex/paper.tex
Add Simplifier section.

@@
 \usepackage{hyperref}
+\usepackage{caption}
 \usepackage[nameinlink, noabbrev]{cleveref}
@@
 They also offer a “Freemium” instance, available by registering via an email address, which has no such limitation, we therefore applied for such an instance.
-%TODO
+Shortly afterwards, we received a confirmation email with the URL to our instance, which unfortunately returned a “400 Bad Request” error.
+We contacted the Simplifier Support Team about this and decided to get familiar with the platform using the Simplifier Playground in the meantime.
 
@@
 
+\subsubsection{Simplifier}
+
+Upon logging into a Simplifier instance, the user sees the Simplifier dashboard, seen in \cref{fig:simplifier_dashboard}.
+The dashboard provides an overview of all elements represented by tiles.
+The “Applications” tile opens a list of applications, allowing to edit them and create new ones.
+“Widgets” allows viewing existing building blocks and also creating new reusable components.
+“Connectors” allow specifying connections to external services, e.g.
+for sending emails or accessing a ReST API.
+The “Business Objects” tile is used to create workflows by assigning connectors, plugins and functions to business objects.
+The “Database Designer” is used for creating database schemas
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{simplifier-dashboard}
+  \caption{Simplifier Dashboard}
+  \label{fig:simplifier_dashboard}
+\end{figure}
+
+For our TODO List application, we first needed to create a connector to a database.
+Simplifier allows creating a connector to common MySQL, PostgreSQL and other databases.
+We opted for a \texttt{SQL} connector using the SQLite backend to avoid having to set up an external database server.
+Next, we opened the “Database Designer” and created a schema for a todo list item, choosing the \texttt{SQL} connector we created as data source.
+The schema can then be edited using an entity-relationship (ER) diagram editor.
+We added a entity called \texttt{task} with fields for \texttt{status}, \texttt{title} and \texttt{body}.
+With a click of a button, we could then deploy the changes in the ER diagram to the database.
+
+Next, we created a new application called \texttt{TODO\_List}.
+We added a custom header bar in order to add a \texttt{add\_task} button for adding new tasks and a list view for displaying the tasks.
+We also added a second screen to our application containing a dialog box for adding a new task.
+We then opened the “Process” tab, which allows creating workflows.
+Workflows are represented as stories, which can be edited using a graphical interface.
+First, we created a workflow that opens the “Add Task dialog” which can be seen in \cref{fig:simplifier_story_overview} and \cref{fig:simplifier_story_ui_action}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{simplifier-story-overview}
+  \caption{Simplifier application process story: If the \texttt{add\_task} button is pressed, execute a \texttt{UIAction}.
+The inner logic of the \texttt{UIAction} is shown in \cref{fig:simplifier_story_ui_action}.}
+  \label{fig:simplifier_story_overview}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{simplifier-story-ui-action}
+  \caption{Inner logic of the UIAction component shown in \cref{fig:simplifier_story_overview}: Open the \texttt{add\_task\_dialog} dialog by setting the \texttt{dialogOpened} property to \texttt{true}.}
+  \label{fig:simplifier_story_ui_action}
+\end{figure}
+
+Unfortunately, when creating a second workflow that is supposed to create a new task in the database from the inputs in the dialog, we received a “Maximum call stack exceeded.” error.
+The Simplifier Playground offers two similar applications to ours, a TODO List using a ReST backend and a shopping list using a SQLite backend.
+We took a closer look at the shopping list application and saw that it defines custom functions for accessing the database.
+Our TODO List application on the other hand uses the “Data Service” connector automatically generated for the database schema we created.
+There were no other demo applications using the “Data Service” connector, so we could not compare our process to an existing working solution.
+Additionally, the TODO List demo application did also experience an error when trying to create new tasks.
+
 \subsubsection{Mendix}

### 2021-01-09 tex/paper.tex
Add Assessment of Low-Code Tools section.

@@
 
-\section{Evaluation of Model-Driven Tools}
-\label{sec:evaluation_of_model_driven_tools}
+\section{Assessment of Low-Code Tools}
+\label{sec:assessment_of_tools}
 
-\section{Findings}
+In this section, we assess whether the evaluated low-code platforms are suitable for serious usage by rating the following aspects:
+
+\begin{itemize}
+  \item ease of use
+  \item customisability
+  \item scalability
+  \item suitability for mission-critical applications
+\end{itemize}
+
+\subsubsection{OSBP (Open Standard Business Platform)}
+
+Given that we were not able to evaluate OSBP, we cannot assess whether applications created using it are scalable.
+The same is true for determining whether it is easy to use and customisable.
+It not working with the latest version of the Eclipse IDE (at the time of writing) also makes it very much unsuitable for mission-critical applications.
+
+\subsubsection{Corteza}
+
+The Corteza-Low-Code platform is spreadsheet based, making is very easy to use.
+Creating a simple TODO List application was possible in less than 10 minutes and should be possible in less than 5 minutes once you are familiar with the platform.
+Apart from choosing the order of predefined building blocks, customising the layout of an application is hindered by the fact that it is spreadsheet based.
+The selection of pre-built blocks is very limited, as is the selection of color schemes.
+In terms of scalabilty, creating complex applications will be difficult using the small selection of building blocks.
+Process management applications which don't rely on many different UI elements, like the included Corteza CRM, can definitely be implemented in a scalable manner.
+Given the overall simplicity of the Corteza platform, it is certainly suitable for creating mission-critical applications.
+
+\subsubsection{Oracle APEX (Application Express)}
+
+Creating an application Oracle APEX requires basic SQL knowledge and the page designer is quite complex.
+Nevertheless, we were able to create our TODO List application in less than two hours by reviewing a sample application.
+It is also possible to completely generate the basic structure of an application using a wizard by importing data from a file, making the initial setup even easier.
+For further development, there is a small learning curve for getting familiar with the interface.
+The default theme offers three different menu types and four colour schemes and there is an option to use a custom theme.
+The page editor offers all basic building blocks in addition to some specialised ones, the ability to execute custom actions on page/button events and advanced options for each component.
+Given the flexibility and advanced options of the page designer, it is definitely possible to create scalable applications which can also be used for mission-critical usage.
+
+\subsubsection{Simplifier}
+
+The Simplifier application editor offers an extensive list of UI components for composing an application with advanced options for each component, which is of course associated with a small learning curve to get familiar with the interface.
+A different theme can be uploaded manually.
+All things considered, this make an application very customisable and the platform itself moderately easy to use.
+Adding custom functionality can be done easily by adding so called “stories” using a graphical editor as well, making it very scalable since these “stories” can then be composed by creating custom events.
+Leaving aside the technical difficulties we had during the creation of our TODO List application, we still think it is suitable for creating mission-critical applications, given the commercial support would most likely be able to sort out these issues.
+
+\subsubsection{Mendix}
+
+% TODO
+
+\section{Findings & Possible Future Work}
 \label{sec:findings}
+
+During the research process for searching low-code tools we could evaluate, we found that there is a very small number of open-source low-code platforms.
+Our assumption is that people in open-source community are most often developers and therefore the need for low-code platforms is simply not there.
+It would be interesting to investigate further into this direction, also including other types of platforms and comparing their presence in the open-source vs.
+the commercial space.
+In general, the commercial low-code platforms in our evaluation also seem to offer a more advanced feature set compared to open-source tools.
+
+% TODO: Difference between open-source/commercial, old/new, etc.
 

### 2021-01-10 tex/paper.tex
Add more findings and conclusion.

@@
   \item customisability
+  \item portability
   \item scalability
@@
 The selection of pre-built blocks is very limited, as is the selection of color schemes.
+Corteza uses a MySQL database in the backend, making at least the stored data very portable.
+The created applications are virtually unportable since they are tightly integrated into the Corteza platform, custom scripts are at least partially portable since they are written in JavaScript.
 In terms of scalabilty, creating complex applications will be difficult using the small selection of building blocks.
@@
 Given the flexibility and advanced options of the page designer, it is definitely possible to create scalable applications which can also be used for mission-critical usage.
+Oracle APEX uses their propietary Oracle Database for the backend and custom logic has to be written in PL/SQL, a propietary programming language also by Oracle.
+This makes an Oracle APEX application completely unportable and vendor lock-in is a definate issue.
 
@@
 Leaving aside the technical difficulties we had during the creation of our TODO List application, we still think it is suitable for creating mission-critical applications, given the commercial support would most likely be able to sort out these issues.
+A Simplifier application relies heavily on “stories”, which are stored in a propietary format and there for not portable.
+The data, however, is highly portable since it is stored on external databases via connectors and many different databases are supported.
 
@@
 
-% TODO
+Mendix Studio offers many building-blocks and widgets to choose from and also offers a marketplace where developers can publish custom ones, making the platform very flexible.
+The color scheme of an application can also be completely customised and the marketplace also offers the ability to use custom themes.
+Due to the fact that the data and building blocks are tightly coupled, it is very easy to create basic application logic.
+Application data can be downloaded from backups in SQLite format and external data sources can be installed form the marketplace, making the data very portable.
+The application itself it tightly coupled to the Mendix platform and not at all portable.
+Given the vast amount of building-blocks available (counting those available from the marketplace), Mendix is certainly suitable for building scalable and mission-critical applications.
 
-\section{Findings & Possible Future Work}
+\section{Findings \& Possible Future Work}
 \label{sec:findings}
@@
 
-% TODO: Difference between open-source/commercial, old/new, etc.
+Another thing that stood out is the user-interface design of the low-code platforms.
+There is a clear trend towards a more streamlined user interface in the more mature products like Oracle APEX and Mendix.
+The user interface of, say Corteza, while simple, is not as thought-through and the Simplifier interface, while functional, looks like a generic UI prototype.
+Whether this makes a platform more usable than another lies of course in the eye of the beholder.
+
+A common point of critique of low-code platforms is the lack of portability.
+According to our evaluation, this point is definitely valid.
+Of the platforms we evaluated, none of the applications are portable; at most, the stored data is easily portable by using a common database backend like MySQL.
+There is no standard definition of what a low-code platform looks like let alone any standard defining low-code building blocks which could be used to make porting an application from one platform to another possible.
+Given that many low-code platforms are commercial products, it is also doubtful if there is any incentive to make it easy for customers to switch platforms.
 
@@
 \label{sec:conclusion}
+
+Taking the assessments in \cref{sec:assessment_of_tools} into account, low-code platforms are a valid alternative to model-driven development.
+Some caveats apply however: By their very nature, low-code platforms will not be as flexible, since with model-driven architecture, developers can take advantage of the full feature set of the software stack and customise all aspects of it, while low-code platforms are limited in the ways they can be extended.
+This is of course highly dependent on the platform, which means one cannot argue for or against low-code in all cases.
+The choice between low-code and model-driven architecture cannot be made without evaluating multiple different low-code platforms and assessing whether they can satisfy the requirements for a particular project, similar to choosing a programming language or framework that is most suitable for a particular task.
 

### 2021-01-10 tex/paper.tex
Fix spelling.

@@
 Simplifier allows creating a connector to common MySQL, PostgreSQL and other databases.
-We opted for a \texttt{SQL} connector using the SQLite backend to avoid having to set up an external database server.
-Next, we opened the “Database Designer” and created a schema for a todo list item, choosing the \texttt{SQL} connector we created as data source.
+We opted for a \texttt{SQL} connector using the SQLite back end to avoid having to set up an external database server.
+Next, we opened the “Database Designer” and created a schema for a TODO list item, choosing the \texttt{SQL} connector we created as data source.
 The schema can then be edited using an entity-relationship (ER) diagram editor.
@@
   \includegraphics[width=\textwidth]{simplifier-story-ui-action}
-  \caption{Inner logic of the UIAction component shown in \cref{fig:simplifier_story_overview}: Open the \texttt{add\_task\_dialog} dialog by setting the \texttt{dialogOpened} property to \texttt{true}.}
+  \caption{Inner logic of the \texttt{UIAction} component shown in \cref{fig:simplifier_story_overview}: Open the \texttt{add\_task\_dialog} dialog by setting the \texttt{dialogOpened} property to \texttt{true}.}
   \label{fig:simplifier_story_ui_action}
@@
 Unfortunately, when creating a second workflow that is supposed to create a new task in the database from the inputs in the dialog, we received a “Maximum call stack exceeded.” error.
-The Simplifier Playground offers two similar applications to ours, a TODO List using a ReST backend and a shopping list using a SQLite backend.
+The Simplifier Playground offers two similar applications to ours, a TODO List using a ReST back end and a shopping list using a SQLite back end.
 We took a closer look at the shopping list application and saw that it defines custom functions for accessing the database.
@@
 The selection of pre-built blocks is very limited, as is the selection of color schemes.
-Corteza uses a MySQL database in the backend, making at least the stored data very portable.
+Corteza uses a MySQL database in the back end, making at least the stored data very portable.
 The created applications are virtually unportable since they are tightly integrated into the Corteza platform, custom scripts are at least partially portable since they are written in JavaScript.
@@
 Given the flexibility and advanced options of the page designer, it is definitely possible to create scalable applications which can also be used for mission-critical usage.
-Oracle APEX uses their propietary Oracle Database for the backend and custom logic has to be written in PL/SQL, a propietary programming language also by Oracle.
+Oracle APEX uses their propietary Oracle Database for the back end and custom logic has to be written in PL/SQL, a propietary programming language also by Oracle.
 This makes an Oracle APEX application completely unportable and vendor lock-in is a definate issue.
@@
 According to our evaluation, this point is definitely valid.
-Of the platforms we evaluated, none of the applications are portable; at most, the stored data is easily portable by using a common database backend like MySQL.
+Of the platforms we evaluated, none of the applications are portable; at most, the stored data is easily portable by using a common database back end like MySQL.
 There is no standard definition of what a low-code platform looks like let alone any standard defining low-code building blocks which could be used to make porting an application from one platform to another possible.

### 2021-01-10 tex/paper.tex
Rewrite abstract.

@@
 
-In this paper we present a survey conducted by comparing low-code tools with model-driven tools for developing software in order to get a better understanding of the advantages and drawbacks of both approaches and whether they are a viable alternative to traditional software development.
+In this paper we present a survey conducted by evaluating various low-code tools in order to get a better understanding of the advantages and drawbacks.
+The evaluation entails creating a test application with each low-code tools and assessing factors such as ease of use and scalability.
+Using the results from this evaluation, we try to determine if and when low-code platforms are a viable alternative or even preferable to traditional software development using a model-driven architecture.
 

### 2021-01-10 tex/paper.tex
Fix spelling.

@@
 Apart from choosing the order of predefined building blocks, customising the layout of an application is hindered by the fact that it is spreadsheet based.
-The selection of pre-built blocks is very limited, as is the selection of color schemes.
+The selection of pre-built blocks is very limited, as is the selection of colour schemes.
 Corteza uses a MySQL database in the back end, making at least the stored data very portable.
 The created applications are virtually unportable since they are tightly integrated into the Corteza platform, custom scripts are at least partially portable since they are written in JavaScript.
-In terms of scalabilty, creating complex applications will be difficult using the small selection of building blocks.
+In terms of scalability, creating complex applications will be difficult using the small selection of building blocks.
 Process management applications which don't rely on many different UI elements, like the included Corteza CRM, can definitely be implemented in a scalable manner.
@@
 Oracle APEX uses their propietary Oracle Database for the back end and custom logic has to be written in PL/SQL, a propietary programming language also by Oracle.
-This makes an Oracle APEX application completely unportable and vendor lock-in is a definate issue.
+This makes an Oracle APEX application completely unportable and vendor lock-in is a definite issue.
 
@@
 Mendix Studio offers many building-blocks and widgets to choose from and also offers a marketplace where developers can publish custom ones, making the platform very flexible.
-The color scheme of an application can also be completely customised and the marketplace also offers the ability to use custom themes.
+The colour scheme of an application can also be completely customised and the marketplace also offers the ability to use custom themes.
 Due to the fact that the data and building blocks are tightly coupled, it is very easy to create basic application logic.

### 2021-01-10 tex/paper.tex
Fix spelling.

@@
 The “Business Objects” tile is used to create workflows by assigning connectors, plugins and functions to business objects.
-The “Database Designer” is used for creating database schemas
+There is also a “Database Designer” which is used for creating database schemas.
 
@@
 For our TODO List application, we first needed to create a connector to a database.
-Simplifier allows creating a connector to common MySQL, PostgreSQL and other databases.
+Simplifier allows creating a connector to MySQL, PostgreSQL and other common databases.
 We opted for a \texttt{SQL} connector using the SQLite back end to avoid having to set up an external database server.

### 2021-01-10 tex/final-presentation.tex
Add final presentation.
Add some slides.

--- 2020-10-19 tex/initial-presentation.tex
@@
 
-\date{\formatdate{20}{10}{2020}}
+\date{\formatdate{12}{01}{2021}}
 
@@
       \item Motivation
+      \item Model-Driven Architecture
       \item Low-Code Architecture
-      \item Model-Driven Architecture
       \item Criticisms
@@
 
+  \begin{frame}{Model-Driven Architecture}
+    \begin{itemize}
+      \item provides a set of guidelines for the structuring of specifications
+      \item code (fully or partially) generated from models, e.g. from UML diagrams
+      \item aimed at developers to speed up development
+    \end{itemize}
+  \end{frame}
+
   \begin{frame}{Low-Code Architecture}
@@
       \item typically aimed at end-users rather than developers
-    \end{itemize}
-  \end{frame}
-
-  \begin{frame}{Model-Driven Architecture}
-    \begin{itemize}
-      \item provides a set of guidelines for the structuring of specifications
-      \item code (fully or partially) generated from models, e.g. from UML diagrams
-      \item aimed at developers to speed up development
     \end{itemize}
@@
 
-  \begin{frame}{Planned Project Procedure}
+  \begin{frame}{Evaluation of Low-Code Tools}
     \begin{itemize}
-      \item Evaluate a set of Low-Code tools.
-      \item Evaluate a set of Model-Driven tools.
-      \item For each tool, determine what it can and can't do.
-      \item Compare the two sets with each other and draw some conclusions.
+      \item Find low-code tools in the following categories:
+          \begin{itemize}
+            \item open-source
+            \item developed by well-known company
+            \item developed by unknown company
+            \item old/well-established platform
+            \item new/unestablished
+          \end{itemize}
+      \item Set up each tool
+      \item Build test application with each tool
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Open Standard Business Platform (OBSP)}
+    \begin{itemize}
+      \item open-source
+      \item plug-in for the Eclipse IDE
+      \item community version of the commercial OS.bee product developed by COMPEX
+      \item latest version over one year old
+      \item does not work with latest version of the Eclipse IDE
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Corteza Low Code}
+    \begin{itemize}
+      \item open-source
+      \item part of the Corteza Project initiated by Crust Technology
+      \item the Corteza Project includes a CRM solution built on top of Corteza~Low~Code, among other things
+      \item web-based platform
+      \item test by signing up with a GitHub or Google account or \\
+            by deploying it locally using Docker
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Oracle APEX (Application Express)}
+    \begin{itemize}
+      \item commercial
+      \item initially released as Oracle Flows in 2000
+      \item web-based platform
+      \item test by signing up for an Oracle Cloud account or \\
+            by requesting an APEX workspace
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Simplifier}
+    \begin{itemize}
+      \item commercial
+      \item initially released by iTiZZiMO in 2012
+      \item web-based platform
+      \item test by using the Simplifier Playground (data is wiped every day) or \\
+            by requesting a Simplifier test instance
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Mendix}
+    \begin{itemize}
+      \item commercial
+      \item founded in 2005 as a subsidiary of Siemens
+      \item web-based platform (Mendix Studio) and Windows application \\
+            (Mendix Studio Pro) with advanced features
+      \item test by signing up for a regular account which allows hosting unlimited applications (with 1GB of memory and 0.5GB of storage per application)
     \end{itemize}

### 2021-01-11 tex/final-presentation.tex
Fix spelling.
Refactor some slides.

@@
     \begin{itemize}
-      \item What can and can't low-code and model-driven tools do?
-      \item Are they a viable alternative to traditional development?
+      \item What are the advantages and disadvantages of low-code tools do?
+      \item Are they a viable alternative to model-driven or traditional development?
       \item When to choose one approach over the other?
@@
       \item provides a set of guidelines for the structuring of specifications
+      \item standardise on models in a given domain to reduce code duplication and speed up development
       \item code (fully or partially) generated from models, e.g. from UML diagrams
-      \item aimed at developers to speed up development
+      \item aimed at developers who have good understanding of underlying programming languages
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Model-Driven Architecture}
+    \begin{itemize}
+      \item Example: Swagger
+        \begin{itemize}
+          \item API specification given in OpenAPI format
+          \item API client is generated for the specified programming language
+          \item support for new languages/frameworks can be added by implementing a new generator
+          \item very easy to provide clients for many languages with virtually no development effort
+        \end{itemize}
     \end{itemize}
@@
     \begin{itemize}
+      \item Model-Driven Architecture
+        \begin{itemize}
+          \item UML diagrams lack details included in the code itself.
+          \item “the Code is the design” - Should models be derived from code instead of code from models?
+        \end{itemize}
       \item Low-Code Architecture
@@
                 i.e. applications which are not controlled by a company's IT department.
-        \end{itemize}
-      \item Model-Driven Architecture
-        \begin{itemize}
-          \item UML diagrams lack details included in the code itself.
-          \item “the Code is the design” - Should models be derived from code instead of code from models?
         \end{itemize}
@@
       \item Set up each tool
-      \item Build test application with each tool
+      \item Build a test application (TODO List) with each tool
     \end{itemize}
@@
 
-  \begin{frame}{Open Standard Business Platform (OBSP)}
+  \begin{frame}{Open Standard Business Platform (OSBP)}
     \begin{itemize}
       \item open-source
-      \item plug-in for the Eclipse IDE
+      \item plug-in for the Eclipse IDE developed since 2016
       \item community version of the commercial OS.bee product developed by COMPEX
@@
       \item open-source
-      \item part of the Corteza Project initiated by Crust Technology
+      \item part of the Corteza Project initiated by Crust Technology in 2019
       \item the Corteza Project includes a CRM solution built on top of Corteza~Low~Code, among other things

### 2021-01-11 tex/final-presentation.tex
Finish presentation.

@@
       \item Criticisms
-      \item Planned Project Procedure
+      \item Evaluation of Low-Code Tools
+      \item Findings \& Future Work
+      \item Conclusion
     \end{itemize}
@@
       \item plug-in for the Eclipse IDE developed since 2016
-      \item community version of the commercial OS.bee product developed by COMPEX
+      \item community version of the commercial OS.bee product \\
+            developed by COMPEX
       \item latest version over one year old
@@
 
+  \begin{frame}{Corteza Low Code}
+    \begin{itemize}
+      \item made for building record-based management applications
+      \item process for building the test application:
+        \begin{enumerate}
+          \item create application namespace
+          \item create module for TODO List records, specifying the necessary fields (status, title, body)
+          \item create page and add a list block linked to the module
+        \end{enumerate}
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}[standout]{Corteza Low Code: Editor}
+    \includegraphics[width=\textwidth]{corteza-low-code-pages}
+  \end{frame}
+
+  \begin{frame}[standout]{Corteza Low Code: Application}
+    \begin{figure}
+      \includegraphics[width=\textwidth]{corteza-low-code-crm-home}
+    \end{figure}
+  \end{frame}
+
   \begin{frame}{Oracle APEX (Application Express)}
@@
     \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Oracle APEX (Application Express)}
+    \begin{itemize}
+      \item process for building the test application:
+        \begin{enumerate}
+          \item create database table for TODO List items \\
+                (requires basic SQL knowledge)
+          \item create new blank application
+          \item add list view to home page and select \\
+                the corresponding database table
+          \item add new form page (dialog style) and select \\
+                the corresponding database table
+          \item add a button to the home page that opens the form page
+          \item add a dynamic action that refreshes the list view \\
+                when the form dialog is closed
+        \end{enumerate}
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}[standout]{Oracle APEX: Page Designer}
+    \includegraphics[width=\textwidth]{oracle-apex-page-designer}
+  \end{frame}
+
+  \begin{frame}[standout]{Oracle APEX: Application}
+    \includegraphics[width=\textwidth]{oracle-apex-todo-list-application}
   \end{frame}
@@
 
+  \begin{frame}{Simplifier}
+    \begin{itemize}
+      \item process for building the test application:
+        \begin{enumerate}
+          \item create database connector (SQLite)
+          \item create database schema for TODO List items and deploy to connector
+          \item add list view and button to home page
+          \item add new page containing a form with input fields and button
+          \item create processes for
+            \begin{itemize}
+              \item loading items into list view
+              \item submitting the form page
+              \item opening the form page with the button
+            \end{itemize}
+        \end{enumerate}
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}[standout]{Simplifier: Process}
+    \includegraphics[width=\textwidth]{simplifier-story-overview}
+  \end{frame}
+
+  \begin{frame}[standout]{Simplifier: Process}
+    \includegraphics[width=\textwidth]{simplifier-story-ui-action}
+  \end{frame}
+
+  \begin{frame}[standout]{Simplifier: Application Editor}
+    \includegraphics[width=\textwidth]{simplifier-application-editor}
+  \end{frame}
+
+  \begin{frame}[standout]{Simplifier: Application}
+    \includegraphics[width=\textwidth]{simplifier-shopping-list-application}
+  \end{frame}
+
   \begin{frame}{Mendix}
@@
   \end{frame}
+
+  \begin{frame}{Mendix}
+    \begin{itemize}
+      \item process for building the test application:
+        \begin{enumerate}
+          \item create blank application
+          \item add list view and button to home page, which prompts the user to select or create a new data source
+          \item create data source for TODO List items
+          \item add new form page linked to the corresponding data source
+          \item add button to home page that opens the form page
+        \end{enumerate}
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}[standout]{Mendix Studio}
+    \includegraphics[width=\textwidth]{mendix-studio}
+  \end{frame}
+
+  \begin{frame}[standout]{Mendix: Application}
+    \includegraphics[width=\textwidth]{mendix-todo-list-application}
+  \end{frame}
+
+  \begin{frame}{Unique Features}
+    \begin{itemize}
+      \item \textbf{Oracle APEX}
+        \begin{itemize}
+          \item Creation of applications from existing data, e.g. CSV files
+        \end{itemize}
+      \item \textbf{Mendix}
+        \begin{itemize}
+          \item AI-assisted wizard for creating custom workflows (microflows)
+          \item publishing as native mobile applications for iOS and Android
+        \end{itemize}
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Assessment of Low-Code Tools}
+    \begin{itemize}
+      \item ease of use
+      \item customisability
+      \item portability
+      \item scalability
+      \item suitability for mission-critical applications
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Assessment of Low-Code Tools}
+    \begin{table}[]
+    \begin{tabular}{l|l|l|l|l|l|}
+    \cline{2-6}
+                                           & \textbf{OSBP} & \textbf{Corteza} & \textbf{APEX}   & \textbf{Simplifier} & \textbf{Mendix} \\ \hline
+    \multicolumn{1}{|l|}{ease of use}      & N/A  & high    & medium & medium     & high   \\ \hline
+    \multicolumn{1}{|l|}{customisability}  & N/A  & low     & medium & medium     & high   \\ \hline
+    \multicolumn{1}{|l|}{portability}      & N/A  & medium  & low    & medium     & medium \\ \hline
+    \multicolumn{1}{|l|}{scalability}      & N/A  & medium\footnote{medium in general, high for certain types of applications, e.g. management applications} & high   & high       & high   \\ \hline
+    \multicolumn{1}{|l|}{mission-criticality} & low  & high    & high   & medium\footnote{medium only due to the problems encountered, high otherwise}    & high   \\ \hline
+    \end{tabular}
+    \end{table}
+  \end{frame}
+
+  \begin{frame}{Findings \& Future Work}
+    \begin{itemize}
+      \item very small number of open-source low-code tools
+        \begin{itemize}
+          \item assumption: open-source community mostly consists of developers, so there is no need/demand for low-code platforms
+          \item future work: investigate low-code other types of platforms and compare their presence in the open-source vs. the commercial space
+        \end{itemize}
+      \item more streamlined user interface in the more mature products like Oracle APEX and Mendix
+      \item lack of portability is a valid concern for low-code tools, \\
+            stored data may be the only element of a platform that is portable by using
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}{Conclusion}
+    \begin{itemize}
+      \item low-code platforms are a valid alternative to model-driven development
+      \item low-code platforms are not as flexible and limited in the ways they can be extended
+      \item choice between low-code and model-driven architecture:
+        \begin{itemize}
+          \item highly dependent on which low-code platform is used, \\
+                similar to choosing a programming language or framework for a particular task
+          \item highly dependent on the application requirements, \\
+                low-code well-suited for management applications (e.g. process management, CRM)
+        \end{itemize}
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}[standout]
+    Questions?
+  \end{frame}
 \end{document}

### 2021-01-11 tex/paper.tex
Add some slides.
Finish presentation.

@@
 Shortly afterwards, we received a confirmation email with the URL to our instance, which unfortunately returned a “400 Bad Request” error.
-We contacted the Simplifier Support Team about this and decided to get familiar with the platform using the Simplifier Playground in the meantime.
+We contacted the Simplifier support team about this and decided to get familiar with the platform using the Simplifier Playground in the meantime.
 
@@
 
-The Corteza-Low-Code platform is spreadsheet based, making is very easy to use.
+The Corteza-Low-Code platform is meant for building record-based management applications, making is very easy to use.
 Creating a simple TODO List application was possible in less than 10 minutes and should be possible in less than 5 minutes once you are familiar with the platform.
-Apart from choosing the order of predefined building blocks, customising the layout of an application is hindered by the fact that it is spreadsheet based.
+Apart from choosing the order of predefined building blocks, customising the layout of an application is hindered by the fact that it is specialised for building management applications.
 The selection of pre-built blocks is very limited, as is the selection of colour schemes.

### 2021-01-12 tex/final-presentation.tex
Tweak formatting.
Refactor presentation.

@@
       \item Evaluation of Low-Code Tools
+      \item Assessment of Low-Code Tools
       \item Findings \& Future Work
@@
   \begin{frame}{Model-Driven Architecture}
-    \begin{itemize}
-      \item Example: Swagger
+    Example: Code Generation using Swagger
+
     \begin{itemize}
@@
       \item API client is generated for the specified programming language
-          \item support for new languages/frameworks can be added by implementing a new generator
-          \item very easy to provide clients for many languages with virtually no development effort
-        \end{itemize}
+      \item support for new languages/frameworks can be added \\
+            by implementing a new generator
+      \item very easy to provide API clients for many languages with \\
+            virtually no development effort
     \end{itemize}
@@
             application logic as well as the user interface
-      \item typically aimed at end-users rather than developers
+      \item typically aimed more at end-users rather than developers
+      \item no-code: targeted completely towards end-users
     \end{itemize}
@@
         \begin{itemize}
-          \item UML diagrams lack details included in the code itself.
+          \item UML diagrams lack details included in the code itself
           \item “the Code is the design” - Should models be derived from code instead of code from models?
@@
         \begin{itemize}
-          \item Unsuitable for implementing scalable and mission-critical applications.
-          \item Increase in unsupported applications built by shadow IT,
-                i.e. applications which are not controlled by a company's IT department.
+          \item unsuitable for implementing scalable and mission-critical applications
+          \item increase in unsupported applications built by shadow IT,
+                i.e. applications which are not controlled by a company's IT department
         \end{itemize}
@@
       \item Set up each tool
-      \item Build a test application (TODO List) with each tool
+      \item Build a test application (TODO List) with each tool \\
+        \begin{itemize}
+          \item add/edit items
+          \item mark items as done
+        \end{itemize}
     \end{itemize}
@@
       \item low-code platforms are a valid alternative to model-driven development
-      \item low-code platforms are not as flexible and limited in the ways they can be extended
+      \item low-code platforms are not as flexible and limited in \\
+            the ways they can be extended
       \item choice between low-code and model-driven architecture:
         \begin{itemize}
-          \item highly dependent on which low-code platform is used, \\
-                similar to choosing a programming language or framework for a particular task
+          \item highly dependent on which low-code platform is used, similar to \\
+                choosing a programming language or framework for a particular task
           \item highly dependent on the application requirements, \\
-                low-code well-suited for management applications (e.g. process management, CRM)
+                low-code well-suited for management applications \\
+                (e.g. process management, CRM)
         \end{itemize}

### 2021-01-17 tex/paper.tex
Add todos.
Proof-read paper.

@@
 
-In this paper we present a survey conducted by evaluating various low-code tools in order to get a better understanding of the advantages and drawbacks.
-The evaluation entails creating a test application with each low-code tools and assessing factors such as ease of use and scalability.
+In this paper we present a survey conducted by evaluating various low-code tools in order to get a better understanding of their advantages and drawbacks.
+The evaluation entails creating a test application with each low-code tool and assessing factors such as ease of use, customisability and portability.
 Using the results from this evaluation, we try to determine if and when low-code platforms are a viable alternative or even preferable to traditional software development using a model-driven architecture.
@@
 
-In this paper, we compare Low-Code Platforms to Model-Driven Architecture and try to determine when and where to use them over Model-Driven Architecture or vice-versa.
-
-In order to make clear to the reader the difference between the two approaches we firstly provide definitions and characteristics of each paradigm in the following sections.
+In this paper, we compare various low-code platforms and try to determine when and where to use them over model-driven architecture or vice-versa.
+In order to make the difference between the two approaches clear to the reader, we firstly provide definitions and characteristics of each paradigm in the following two subsections.
+Afterwards, \cref{sec:selection_of_low_code_tools} will give an overview of the low-code tools that were chosen for the evaluation and \cref{sec:developing_a_todo_list_application} then describes the creation of a test application with each platform.
+\Cref{sec:assessment_of_tools} then shows an assessment of each tool based on various criteria which is used as the basis for the findings in \cref{sec:findings}.
+Finally, we present a conclusion of the results in \cref{sec:conclusion}.
 
@@
 
-With Model-Driven Architecture or Model-Driven Engineering, the goal is to standardise on models in a given domain.
+With model-driven architecture or model-driven engineering, the goal is to standardise on models in a given domain.
 This is done in order to reduce code duplication and speed up development.
-Especially when bootstrapping a new project, it is helpful keep boilerplate code to a minimum.
-In addition to reducing development time, Model-Driven Engineering also tries to increase productivity by maximising compatibility between systems by using the aforementioned standardised models.~\cite{wiki:model_driven_engineering} Generally, Model-Driven Architectures are geared towards developers, i.e.
+Especially when bootstrapping a new project, it is helpful to keep boilerplate code to a minimum.
+In addition to reducing development time, model-driven engineering also tries to increase productivity by maximising compatibility between systems by using the aforementioned standardised models.~\cite{wiki:model_driven_engineering} Generally, model-driven architectures are geared towards developers, i.e.
 people who also have a good understanding of the underlying programming language(s) of the respective architecture.
-One common way Model-Driven Engineering is used in practice is code generation.
+One common way model-driven engineering is used in practice is code generation.
 Using code generation can speed up many processes: For example, given an API specification (e.g.
-using the industry standard OpenAPI~\cite{openapis}), one can generate all code that is necessary for a client for this API.
+using the industry standard OpenAPI~\cite{openapis} format), one can generate all code that is necessary for a client for this API.
 Swagger~\cite{swagger} is one such tool which can generate API clients for many different programming languages.
 There are also tools which can generate a complete class hierarchy from an UML diagram; many IDEs offer such functionality either natively or via plug-ins.
-These are only a few examples of what falls into the category of Model-Driven Architecture, the main point is that all of them help developers reach their goal faster than doing the same tasks manually.
+These are only a few examples of what falls into the category of model-driven architecture, the main point is that all of them help developers reach their goal faster than doing the same tasks manually.
 
-\subsection{What is a Low-Code Platform?}
+\subsection{What is a Low-Code Architecture?}
 \label{ssec:what_is_a_low_code_platform}
 
-A low-code platform, as the name implies, requires a minimal amount of code to be written.
+A low-code architecture, as the name implies, requires minimal amount of code to be written.
 Similarly to conventional programming (i.e.
 using a programming language), an IDE is used for creating programs using low-code platforms.
-Often, the IDE is part of the low-code platform itself instead of a standalone program as is the case for conventional programming.
-In most cases, the IDE of the platform and the low-code platform cannot be used separately so the term low-code platform is also commonly used to refer to the IDE interface itself.
+Often, the IDE is part of the low-code platform itself instead of a standalone program as is the case for conventional programming.~\cite{wiki:low_code_development_platform} Furthermore, the IDE of a low-code platform is often tightly integrated into the platform, so the term low-code platform is also commonly used to refer to the interface itself.
 Low-code platforms are targeted towards users with little to no programming experience, which means the actual programming of a low-code system is done mainly via visual building blocks.
@@
 Oftentimes, custom functionality can be implemented using a click-and-drag pattern between building blocks and by assigning an action accordingly.
-The available building blocks vary by platform which makes them less portable than a system written using a programming language.
-This is one common point of critique on low-code platforms.
+The available building blocks vary from platform to platform, making them less portable than a system written using a programming language.
+This is one common point of critique on low-code platforms.~\cite{low_code_platforms_deliver_customer_facing_apps_but_will_they_scale_up, low_code_patforms_dont_buy_the_hype, the_benefits_of_implementing_low_code_development_platforms}
 
-Another sub-category of low-code platforms are so called no-code platforms~\cite{wiki:no_code_development_platform}.
+Another subcategory of low-code platforms are so called no-code platforms~\cite{wiki:no_code_development_platform}.
 Where to draw the line between low-code and no-code is not clearly defined, especially since low-code platforms can sometimes be used in a way that cannot be differentiated from no-code platforms.
-For the purposes of this paper, low-code platforms are general purpose visual programming languages~\cite{wiki:visual_programming_language} aimed at both developer and non-developer users while no-code platforms are closer to software as a service (SaaS) platforms targeted mostly (or only) towards non-developers.
-Another way to look at it is that low-code platforms still give developers the freedom to add custom building blocks for their specific use case and no-code solutions are usually purpose-made for a specific line of business or a specific type of department.
+For the purposes of this paper, low-code platforms are general purpose visual programming languages~\cite{wiki:visual_programming_language} aimed at both developer and non-developer users while no-code platforms are closer to software as a service (SaaS~\cite{wiki:saas}) platforms targeted mostly (or only) towards non-developers.
+Another way to look at it is that low-code platforms still give developers the freedom to add custom building blocks for their specific use case while no-code solutions are usually purpose-made for a specific line of business or a specific type of department.
 Some examples include monday.com~\cite{wiki:monday_com}, a no-code platform for project management tasks; FileMaker~\cite{wiki:filemaker}, which provides a database engine and a corresponding no-code platform for creating graphical user interfaces for interacting with databases; and Airtable~\cite{wiki:airtable}, a no-code platform for creating databases connected to interactive spreadsheets.
 
-\section{Evaluation of Low-Code Tools}
-\label{sec:evaluation_of_low_code_tools}
+\section{Selection of Low-Code Tools}
+\label{subsec:selection_of_low_code_tools}
 
-\subsection{Selection of Low-Code Tools}
-\label{subsec:choosing_low_code_tools}
-
-Before evaluating low-code tools, we had to first determine which tools to evaluate.
+Before evaluating low-code tools, we had to first determine which ones to evaluate.
 Given the ever evolving landscape of low-code platforms, there are countless companies offering low-code solutions.
@@
 Simplifier or Corteza).
-With the given list of low-code platforms, we could satisfy this requirement.
+With the platforms listed above, we could satisfy this requirement.
 
-To ensure we could properly test all of these tools, we conducted a brief first inspection of each.
-Therefore, the following section will give a brief explanation of each platform, any problems encountered during the initial setup and the steps necessary before being able to use the low-code platform.
+To ensure we could properly test all of these tools, we conducted an initial inspection of each.
+Therefore, the rest of this subsection will give a brief explanation of each platform, any problems encountered during the initial setup and the steps necessary before being able to use the low-code platform.
 
-\subsubsection{OSBP (Open Standard Business Platform)}
+\subsection{OSBP (Open Standard Business Platform)}
 
 The Open Standard Business Platform is an open-source development environment which combines no-code and low-code with traditional application development.~\cite{osbp} It is the community version of OS.bee~\cite{osbee}, a commercial product developed by COMPEX only available to business users.
-The interface for the OSBP platform is implemented via a plug-in for the Eclipse IDE.
+The interface for the OSBP platform is implemented via a plug-in for the Eclipse IDE~\cite{eclipse_ide}.
 The user first needs to install the Eclipse IDE, after which they can add the link to the plug-in repository and install it directly from within the IDE.
-Unfortunately, the latest version of OSBP was released more than one year ago and is incompatible with the latest version of the Eclipse IDE.
-At the time of writing (Jan 2, 2021), the last commit in the OSBP Git repository was on May 9, 2019.
-The latest version of the Eclipse IDE was 2020-12.
-The OSBP website recommended downloading Eclipse Neon, a version of the Eclipse IDE released in 2016.
+Unfortunately, at the time of writing (Jan 2, 2021), the latest version of OSBP was released more than one year ago (May 9, 2019) and is incompatible with the latest version of the Eclipse IDE (Dec 2020).
+The OSBP website recommends downloading Eclipse Neon, a version of the Eclipse IDE released in 2016.
 A non-functioning IDE plug-in for an outdated version of said IDE which has not been updated in over a year is a red flag.
@@
 
-\subsubsection{Corteza}
+\subsection{Corteza}
 
@@
 It is possible to try out Corteza directly in the browser by creating an account or by using an existing GitHub or Google account to sign up.
-We chose to set up a local demo deployment using Docker.
+We chose the alternative option of setting up a local demo deployment using Docker.
 The Corteza Project provides very good documentation on how to get started, including a ready-made \texttt{docker-compose.yml} file containing all necessary services for a full Corteza installation.
 After a simple call to \texttt{docker-compose up -d}, the Corteza server was running locally.
-The last step before being ready to use was to set up an administrator account in the local Corteza instance.
+The last step before being ready to use it was to set up an administrator account in the local Corteza instance.
 
-\subsubsection{Oracle APEX (Application Express)}
+\subsection{Oracle APEX (Application Express)}
 
@@
 For our purposes, we chose to request a free workspace for testing.
-We chose a name for our namespace and registered using an email address and shortly thereafter we were sent an approval email with the login details.
+We chose a name for our namespace and registered using an email address and shortly thereafter were sent an approval email with the login details.
 After that, Oracle APEX was ready to use.
 
-\subsubsection{Simplifier}
+\subsection{Simplifier}
 
 Simplifier is, like Oracle APEX, a web-based application development environment on low-code basis.
-Simplifier offers a demo platform called “Simplifier Playground”, an instance of the Simplifier platform specifically set up for trialling the platform.
-The trial instance is reset to defaults every night, so all data stored on it will not be persisted.
-They also offer a “Freemium” instance, available by registering via an email address, which has no such limitation, we therefore applied for such an instance.
-Shortly afterwards, we received a confirmation email with the URL to our instance, which unfortunately returned a “400 Bad Request” error.
+Simplifier offers a demo platform called Simplifier Playground, an instance of the Simplifier platform specifically set up for trialling the platform.
+The trial instance is reset every night, so all data stored on it is not persisted.
+They also offer a “Freemium” instance, available by registering via an email address.
+We therefore applied for such an instance since it has no such limitation.
+Shortly afterwards, we received a confirmation email with the URL to our instance, which unfortunately returned a “400~Bad~Request” error.
 We contacted the Simplifier support team about this and decided to get familiar with the platform using the Simplifier Playground in the meantime.
 
-\subsubsection{Mendix}
+\subsection{Mendix}
 
@@
 There are two options for developing applications using Mendix: The first is the web-based environment called Mendix Studio, the second a native desktop application called Mendix Studio Pro, offering more advanced options for professional users.
-Unfortunately Mendix Studio Pro is only available for Microsoft Windows, so we were not able to test it, the basic functionality for creating low-code application is the same in both versions of Mendix Studio, however.
+Unfortunately Mendix Studio Pro is only available for Microsoft Windows, we were therefore not able to test it, the basic functionality for creating low-code application is the same in both versions of Mendix Studio, however.
 
-\subsection{Developing a TODO List Application}
+\section{Developing a TODO List Application}
+\label{sec:developing_a_todo_list_application}
 
 In order to evaluate each platform, we decided to build a simple TODO list application.
-The user should be able to add items to the list containing a title, body text and the status of the TODO item (either “done” or “outstanding”).
+The user should be able to add TODO items to a list containing a title, body text and the status of the TODO item (either “done” or “outstanding”).
+The following subsections describe the creation process of such an application using each platform.
 
-\subsubsection{Corteza}
+\subsection{Corteza}
 
@@
 By selecting “Low Code”, we reach a new view containing all existing “Low Code Namespaces”.
-A namespace can be thought of as an individual application within the low-code environment.
+A namespace in Corteza can be thought of as an individual application within the low-code environment.
 In order to start with our application, we create a \texttt{todo-list} namespace.
@@
 Without this option, the user has to click on the list item to open the detail page in order to edit it.
-Unfortunately, for some reason when enabling the checkbox to allow inline editing, we were unable to save the page with this configuration.
+Unfortunately, when enabling the checkbox to allow inline editing, we were unable to save the page with this configuration for some reason.
 Nevertheless, the basic functionality of our application was completely implemented in less than ten minutes.
@@
 
-\subsubsection{Oracle APEX (Application Express)}
+\subsection{Oracle APEX (Application Express)}
 
@@
 
-\subsubsection{Simplifier}
+\subsection{Simplifier}
 
@@
 We opted for a \texttt{SQL} connector using the SQLite back end to avoid having to set up an external database server.
-Next, we opened the “Database Designer” and created a schema for a TODO list item, choosing the \texttt{SQL} connector we created as data source.
+Next, we opened the “Database Designer” and created a schema for TODO list items, choosing the \texttt{SQL} connector we created as data source.
 The schema can then be edited using an entity-relationship (ER) diagram editor.
@@
 We added a custom header bar in order to add a \texttt{add\_task} button for adding new tasks and a list view for displaying the tasks.
-We also added a second screen to our application containing a dialog box for adding a new task.
+We also added a second screen to our application containing a dialog box for adding new tasks.
 We then opened the “Process” tab, which allows creating workflows.
-Workflows are represented as stories, which can be edited using a graphical interface.
-First, we created a workflow that opens the “Add Task dialog” which can be seen in \cref{fig:simplifier_story_overview} and \cref{fig:simplifier_story_ui_action}.
+Workflows are represented as so-called “stories” which can be edited using a graphical interface.
+First, we created a workflow that opens the \texttt{add\_task} dialog which can be seen in \cref{fig:simplifier_story_overview} and \cref{fig:simplifier_story_ui_action}.
 
@@
 Unfortunately, when creating a second workflow that is supposed to create a new task in the database from the inputs in the dialog, we received a “Maximum call stack exceeded.” error.
-The Simplifier Playground offers two similar applications to ours, a TODO List using a ReST back end and a shopping list using a SQLite back end.
-We took a closer look at the shopping list application and saw that it defines custom functions for accessing the database.
-Our TODO List application on the other hand uses the “Data Service” connector automatically generated for the database schema we created.
+The Simplifier Playground offers two demo applications similar to ours, a TODO List using a ReST back end and a shopping list using a SQLite back end.
+We took a closer look at the shopping list application and saw that it defines custom functions for accessing the database using JavaScript.
+Our TODO List application on the other hand uses the “Data Service” connector automatically generated from the database schema we created.
 There were no other demo applications using the “Data Service” connector, so we could not compare our process to an existing working solution.
@@
 
-\subsubsection{Mendix}
+\subsection{Mendix}
 
@@
 We developed our TODO list application entirely using the web interface.
-The Mendix interface provides four main menu items: Pages, Domain Models, Microflows and Navigation.
+The Mendix Studio interface provides four main menu items: Pages, Domain Models, Microflows and Navigation.
 Pages is used to manage the different pages needed for the application.
@@
 when clicking a button.
-An interesting feature is Mendix Assist, an AI-based assistant making suggestions to the user when creating Microflows.
+An interesting feature is Mendix Assist, an AI-based assistant making suggestions to the user when creating microflows.
 Lastly we have Navigation, which is simply the representation of the application menu hierarchy.
@@
 There are numerous page templates to choose from, we used the “Form Horizontal” template as seen in \cref{fig:mendix_create_page}.
-The resulting page (\texttt{add\_task}) contains a so called data view.
+The resulting \texttt{add\_task} page contains a so called data view.
 Every data view needs a data source, so we were prompted to create a new entity.
@@
 In order to actually add new tasks, we also added a “New task” button.
-For the click action, the user can choose to open a page or run a Microflow or a predefined action such as saving changes on the current page.
-We chose our page for creating a new task as the target and also needed to enable the “Create object” option.
+For the click action, the user can choose to open a page, run a microflow or a predefined action such as saving changes on the current page.
+We chose our \texttt{add\_task} page as the target and also needed to enable the “Create object” option.
 With this option, we pass an empty \texttt{task} object to the page, since the data view on the page expects a \texttt{task} as data source.
 
-Lastly, we created another form page (\texttt{edit\_task}), also containing the \texttt{status}, in order to be able to edit a \texttt{task}.
+Lastly, we created another form page (\texttt{edit\_task}), also containing the \texttt{status}, in order to be able to make a \texttt{task} as “done”.
 To access it, we added a click event to the list view on the \texttt{home} page.
@@
 \begin{itemize}
-  \item ease of use
-  \item customisability
-  \item portability
-  \item scalability
+  \item ease of use considering setup time, interface complexity and development time
+  \item customisability of themes, layout and building blocks
+  \item portability of data and application itself
+  \item scalability of application functionality
   \item suitability for mission-critical applications
@@
 
-\subsubsection{OSBP (Open Standard Business Platform)}
+\begin{table}[]
+  \centering
+  \begin{tabular}{l|l|l|l|l|l|}
+    \cline{2-6}                               & \textbf{OSBP} & \textbf{Corteza} & \textbf{APEX}   & \textbf{Simplifier} & \textbf{Mendix} \\ \hline
+    \multicolumn{1}{|l|}{ease of use}         & N/A  & high    & medium & medium     & high   \\ \hline
+    \multicolumn{1}{|l|}{customisability}     & N/A  & low     & medium & medium     & high   \\ \hline
+    \multicolumn{1}{|l|}{portability}         & N/A  & medium  & low    & medium     & medium \\ \hline
+    \multicolumn{1}{|l|}{scalability}         & N/A  & medium  & high   & high       & high   \\ \hline
+    \multicolumn{1}{|l|}{mission-criticality} & low  & high    & high   & medium     & high   \\ \hline
+  \end{tabular}
+  \caption{Assessment Overview}
+  \label{tab::assessment}
+\end{table}
 
-Given that we were not able to evaluate OSBP, we cannot assess whether applications created using it are scalable.
-The same is true for determining whether it is easy to use and customisable.
-It not working with the latest version of the Eclipse IDE (at the time of writing) also makes it very much unsuitable for mission-critical applications.
+\Cref{tab::assessment} shows an overview of the assessed criteria rated either low, medium or high. The rationale for this rating is explained for each platform in the following subsections.
 
-\subsubsection{Corteza}
+\subsection{OSBP (Open Standard Business Platform)}
+
+Given that we were not able to evaluate OSBP, we cannot assess whether it is easy to use or whether applications created using it are customisable, portable or scalable.
+It not working with the latest version of the Eclipse IDE (at the time of writing) makes it very much unsuitable for mission-critical applications.
+
+\subsection{Corteza}
 
 The Corteza-Low-Code platform is meant for building record-based management applications, making is very easy to use.
-Creating a simple TODO List application was possible in less than 10 minutes and should be possible in less than 5 minutes once you are familiar with the platform.
+The setup time was below 10 minutes and creating a simple TODO List application was also possible in less than 10 minutes and should be possible in less than 5 minutes once you are familiar with the platform.
 Apart from choosing the order of predefined building blocks, customising the layout of an application is hindered by the fact that it is specialised for building management applications.
-The selection of pre-built blocks is very limited, as is the selection of colour schemes.
+The selection of pre-built blocks is also very limited, as is the selection of colour schemes, making it not very customisable.
 Corteza uses a MySQL database in the back end, making at least the stored data very portable.
-The created applications are virtually unportable since they are tightly integrated into the Corteza platform, custom scripts are at least partially portable since they are written in JavaScript.
+The created applications are virtually unportable since they are tightly integrated into the Corteza platform.
+Custom scripts are at least partially portable since they are written in JavaScript.
 In terms of scalability, creating complex applications will be difficult using the small selection of building blocks.
-Process management applications which don't rely on many different UI elements, like the included Corteza CRM, can definitely be implemented in a scalable manner.
+Process management applications which don't rely on many different UI elements, like the included Corteza CRM Suite, can definitely be implemented in a scalable manner.
 Given the overall simplicity of the Corteza platform, it is certainly suitable for creating mission-critical applications.
 
-\subsubsection{Oracle APEX (Application Express)}
+\subsection{Oracle APEX (Application Express)}
 
-Creating an application Oracle APEX requires basic SQL knowledge and the page designer is quite complex.
+Creating an application with Oracle APEX from scratch requires basic SQL knowledge and the page designer is quite complex.
 Nevertheless, we were able to create our TODO List application in less than two hours by reviewing a sample application.
@@
 The page editor offers all basic building blocks in addition to some specialised ones, the ability to execute custom actions on page/button events and advanced options for each component.
-Given the flexibility and advanced options of the page designer, it is definitely possible to create scalable applications which can also be used for mission-critical usage.
 Oracle APEX uses their propietary Oracle Database for the back end and custom logic has to be written in PL/SQL, a propietary programming language also by Oracle.
 This makes an Oracle APEX application completely unportable and vendor lock-in is a definite issue.
+Given the flexibility and advanced options of the page designer, it is definitely possible to create scalable applications which can also be used for mission-critical usage.
 
-\subsubsection{Simplifier}
+\subsection{Simplifier}
 
 The Simplifier application editor offers an extensive list of UI components for composing an application with advanced options for each component, which is of course associated with a small learning curve to get familiar with the interface.
-A different theme can be uploaded manually.
+Adding custom functionality can be done easily by adding so called “stories” using a graphical editor as well, and different theme can be uploaded manually if needed.
 All things considered, this make an application very customisable and the platform itself moderately easy to use.
-Adding custom functionality can be done easily by adding so called “stories” using a graphical editor as well, making it very scalable since these “stories” can then be composed by creating custom events.
-Leaving aside the technical difficulties we had during the creation of our TODO List application, we still think it is suitable for creating mission-critical applications, given the commercial support would most likely be able to sort out these issues.
-A Simplifier application relies heavily on “stories”, which are stored in a propietary format and there for not portable.
+The “stories” can also be composed by creating custom events, which means custom functionality can be built in a scalable manner.
+Leaving aside the technical difficulties we had during the creation of our TODO List application, we still think it is suitable for creating mission-critical applications, given the commercial support would most likely be able to resolve these issues.
+A Simplifier application relies heavily on the aforementioned “stories”, which are stored in a propietary format and therefore not portable.
 The data, however, is highly portable since it is stored on external databases via connectors and many different databases are supported.
 
-\subsubsection{Mendix}
+\subsection{Mendix}
 
 Mendix Studio offers many building-blocks and widgets to choose from and also offers a marketplace where developers can publish custom ones, making the platform very flexible.
-The colour scheme of an application can also be completely customised and the marketplace also offers the ability to use custom themes.
-Due to the fact that the data and building blocks are tightly coupled, it is very easy to create basic application logic.
-Application data can be downloaded from backups in SQLite format and external data sources can be installed form the marketplace, making the data very portable.
+The colour scheme of an application can also be completely customised and the marketplace also offers the ability to use custom themes, making applications very customisable.
+Due to the fact that the data and building blocks are tightly coupled in the interface, it is very easy to create basic application logic.
+The AI-based microflows wizard also simplifies the creation of more complex logic.
+Application data can be downloaded from backups in SQLite format and external data sources can be installed from the marketplace, making the data very portable.
 The application itself it tightly coupled to the Mendix platform and not at all portable.
@@
 During the research process for searching low-code tools we could evaluate, we found that there is a very small number of open-source low-code platforms.
-Our assumption is that people in open-source community are most often developers and therefore the need for low-code platforms is simply not there.
+Our assumption is that people in the open-source community are most often developers and therefore the need for low-code platforms is simply not there.
 It would be interesting to investigate further into this direction, also including other types of platforms and comparing their presence in the open-source vs.
@@
 There is a clear trend towards a more streamlined user interface in the more mature products like Oracle APEX and Mendix.
-The user interface of, say Corteza, while simple, is not as thought-through and the Simplifier interface, while functional, looks like a generic UI prototype.
-Whether this makes a platform more usable than another lies of course in the eye of the beholder.
+The user interface of, say Corteza, while simple, is not as thought-through and the Simplifier interface, while functional, looks like a generic UI prototype and generally not as polished.
+Whether this makes a platform more usable than another may of course lie in the eye of the beholder and would certainly be worth investigating in more detail.
 
-A common point of critique of low-code platforms is the lack of portability.
-According to our evaluation, this point is definitely valid.
-Of the platforms we evaluated, none of the applications are portable; at most, the stored data is easily portable by using a common database back end like MySQL.
+A common point of critique of low-code platforms is the lack of portability.~\cite{low_code_platforms_deliver_customer_facing_apps_but_will_they_scale_up, low_code_patforms_dont_buy_the_hype, the_benefits_of_implementing_low_code_development_platforms} According to our evaluation, this point is definitely valid.
+Of the platforms we evaluated, none of the applications are portable; at most the stored data is portable if a common database back end like MySQL is used.
 There is no standard definition of what a low-code platform looks like let alone any standard defining low-code building blocks which could be used to make porting an application from one platform to another possible.
@@
 
-Taking the assessments in \cref{sec:assessment_of_tools} into account, low-code platforms are a valid alternative to model-driven development.
-Some caveats apply however: By their very nature, low-code platforms will not be as flexible, since with model-driven architecture, developers can take advantage of the full feature set of the software stack and customise all aspects of it, while low-code platforms are limited in the ways they can be extended.
-This is of course highly dependent on the platform, which means one cannot argue for or against low-code in all cases.
-The choice between low-code and model-driven architecture cannot be made without evaluating multiple different low-code platforms and assessing whether they can satisfy the requirements for a particular project, similar to choosing a programming language or framework that is most suitable for a particular task.
+Taking the assessment in \cref{sec:assessment_of_tools} into account, low-code platforms are a valid alternative to model-driven development.
+Some caveats apply however: By their very nature, low-code platforms will not be as flexible, since with model-driven architecture developers can take advantage of the full feature set of the chosen software stack and customise all aspects of it, while low-code platforms are limited in the ways they can be extended.
+This is of course highly dependent on the platform, which means one cannot argue for or against low-code in a general sense.
+The choice between low-code and model-driven architecture cannot be made without evaluating multiple different low-code platforms and assessing whether they can satisfy the requirements for a particular project, similar to choosing a specific programming language or framework that is most suitable for a particular task.
 
